---
title: "Analyze external datasets"
author: "Drake Williams"
date: "`r BiocStyle::doc_date()`"
output:   
  pdf_document:
    dev: cairo_pdf
knit: (function(inputFile, encoding) { 
    rmarkdown::render(
        inputFile, encoding = encoding, 
        output_file = file.path(
            dirname(inputFile), paste0('05_02_ExternalDataAnalysis_',Sys.Date(),'.html'))) 
                })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = F,
  message = F, 
  out.width = "100%",
  fig.align = "center",
  time_it = T
)
options(width = 1200)
```

# Introduction

The objective of this notebook is to load datasets from CZI CELLxGENE and change ENSEMBL gene names to gene symbols

## Load packages

```{r load packages}
# load packages
library(pacman)
p_load(Seurat, 
       scCustomize, 
       dplyr, 
       org.Hs.eg.db, 
       ggpubr,
       data.table,
       paletteer,
       Cairo,
       rstatix)
source("~/epithelial-IL23/01_scripts/analysisHelpers.R")
```
# Import all CZI datasets
```{r generate czi seurat list}
folder <- "/home/williamsdrw/epithelial-IL23/02_raw_data/czi/"
seurat_objects <- read_rds_files(folder)
```

## Modify metadata and trim CZI datasets
### Yoshida 2022
```{r y, fig.width=5.5, fig.height=3}
# modify metadata
seurat_objects[[1]]$ref <- "Yoshida 2022"
seurat_objects[[1]]$source <- "czi_cellxgene"
meta <- seurat_objects[[1]]$sample_id
meta <- gsub("_", "-", as.character(meta))
seurat_objects[[1]]$orig.ident <- meta

# extract epithelial cells
Idents(seurat_objects[[1]]) <- "Cell_type_annotation_level1"
seurat_objects[[1]] <- subset(seurat_objects[[1]], idents = "Epi")
# define the number of total cells to downsample by
# seurat_objects[[1]] <- downsampleSeurat(seurat_objects[[1]], indx)
seurat_objects[[1]] <- convertEnsemblRownames(seurat_objects[[1]])

# generate a seurat object with bulk expression based on cell type, disease status, and donor
bulk <- getBulkExpression(seu = seurat_objects[[1]],
                          cellType = "cell_type",
                          donor = "orig.ident",
                          disease = "disease",
                          diseaseLevels = c("COVID-19", "normal"))
Idents(bulk) <- "celltype"
bulk <- subset(bulk,
               idents = c("neuroendocrine cell",
                          "melanocyte",
                          "duct epithelial cell",
                          "native cell"),
               invert = T)

lungYosh <- VlnPlot_scCustom(bulk, features = c("IL23A"),
                 pt.size = 0.5,
                 colors_use = c("Red", "Gray"),
                 group.by = "celltype",
                 split.by = "disease") +
stat_compare_means(method = "t.test",
                   label = "p.signif",
                     label.y = 6.75) +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.5),
        legend.position = "none") +
  scale_y_continuous(breaks=seq(0,8,2),
                     limits=c(0, 7.25)) +
  coord_flip()

# cell types on y axis: fig.width=5.5, fig.height=7
cairo_pdf(filename="/home/williamsdrw/epithelial-IL23/08_final_plots_for_figures/IL23A-Yoshida2022-lungCOVID.pdf",
          width=4,
          height=3)
lungYosh
dev.off()

yoshStat <- ggplot_build(lungYosh)
```

### Chan 2021
```{r c, fig.width=6.5, fig.height=4.5, eval = F, echo=F}
seurat_objects[[2]]$ref <- "Chan 2021"
seurat_objects[[2]]$source <- "czi_cellxgene"
meta <- seurat_objects[[2]]$donor_id
meta <- gsub("_", "-", as.character(meta))
seurat_objects[[2]]$orig.ident <- meta
# extract epithelial cells
Idents(seurat_objects[[2]]) <- "cell_type"
seurat_objects[[2]] <- subset(seurat_objects[[2]], idents = c("ionocyte", "hepatocyte", "neuroendocrine cell"), invert = T)

# seurat_objects[[2]] <- downsampleSeurat(seurat_objects[[2]], indx)
seurat_objects[[2]] <- convertEnsemblRownames(seurat_objects[[2]])

# generate a seurat object with bulk expression based on cell type, disease status, and donor
bulk <- getBulkExpression(seu = seurat_objects[[2]],
                          cellType = "cell_type",
                          donor = "orig.ident",
                          disease = "disease",
                          diseaseLevels = c("normal", "lung adenocarcinoma", "small cell lung carcinoma"))

# cell types on x axis: fig.width=6.5, fig.height=4.5
VlnPlot_scCustom(bulk, features = c("IL23A"),
                 pt.size = 0.5,
                 colors_use = c("Grey", "Red"),
                 group.by = "celltype",
                 split.by = "disease",
                 y.max = 8) +
  stat_compare_means(method = "anova",
                     label = "p.signif") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45))
# cell types on y axis: fig.width=6.5, fig.height=4.5
VlnPlot_scCustom(bulk, features = c("IL23A"),
                 pt.size = 0.5,
                 colors_use = c("Grey", "Red"),
                 group.by = "celltype",
                 split.by = "disease",
                 y.max = 8) +
  stat_compare_means(method = "anova",
                     label = "p.signif") +
  theme(axis.title.y = element_blank()) +
  coord_flip()
```

### Kong 2023 - Colon
```{r k, fig.width=4.5, fig.height=3}
seurat_objects[[3]]$ref <- "Kong 2023"
seurat_objects[[3]]$source <- "czi_cellxgene"
meta <- seurat_objects[[3]]$biosample_id
meta <- gsub("_", "-", as.character(meta))
seurat_objects[[3]]$orig.ident <- meta

# seurat_objects[[3]] <- downsampleSeurat(seurat_objects[[3]], indx)
seurat_objects[[3]] <- convertEnsemblRownames(seurat_objects[[3]])

# generate a seurat object with bulk expression based on cell type, disease status, and donor
bulk <- getBulkExpression(seu = seurat_objects[[3]],
                          cellType = "cell_type",
                          donor = "orig.ident",
                          disease = "disease",
                          diseaseLevels = c("Crohn disease", "normal"))

# # cell types on x axis: fig.width=6.5, fig.height=4.5
# VlnPlot_scCustom(bulk, features = c("IL23A"),
#                  pt.size = 0.5,
#                  colors_use = c("Grey", "Red"),
#                  group.by = "celltype",
#                  split.by = "disease",
#                  y.max = 3) +
#   stat_compare_means(method = "anova",
#                      label = "p.signif") +
#   theme(axis.title.x = element_blank(),
#         axis.text.x = element_text(angle = 45))
# cell types on y axis: fig.width=6.5, fig.height=4.5
colon <- VlnPlot_scCustom(bulk, features = c("IL23A"),
                 pt.size = 0.5,
                 colors_use = c("Red", "Gray"),
                 group.by = "celltype",
                 split.by = "disease",
                 y.max = 6) +
  stat_compare_means(method = "t.test",
                     label = "p.signif",
                     label.y = 5.75) +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.5)) +
  coord_flip()

# cell types on y axis: fig.width=5.5, fig.height=7
cairo_pdf(filename="/home/williamsdrw/epithelial-IL23/08_final_plots_for_figures/IL23A-Kong2023-colon.pdf",
          width=4.5,
          height=3)
colon  
dev.off()

```
### Sikkema 2023
```{r s, fig.width=7, fig.height=3}
seurat_objects[[4]]$ref <- "Sikkema 2023"
seurat_objects[[4]]$source <- "czi_cellxgene"
meta <- seurat_objects[[4]]$sample
meta <- gsub("_", "-", as.character(meta))
seurat_objects[[4]]$orig.ident <- meta
# extract epithelial cells
Idents(seurat_objects[[4]]) <- "ann_coarse"
seurat_objects[[4]] <- subset(seurat_objects[[4]], idents = c("Epithelial cell"))

# seurat_objects[[4]] <- downsampleSeurat(seurat_objects[[4]], indx)
seurat_objects[[4]] <- convertEnsemblRownames(seurat_objects[[4]])

# generate a seurat object with bulk expression based on cell type, disease status, and donor
bulk <- getBulkExpression(seu = seurat_objects[[4]],
                          cellType = "cell_type",
                          donor = "orig.ident",
                          disease = "disease",
                          diseaseLevels = c("chronic obstructive pulmonary disease",
                                            "lung adenocarcinoma",
                                            "non-small cell lung carcinoma",
                                            "squamous cell lung carcinoma",
                                            "normal"))

# # cell types on x axis: fig.width=6.5, fig.height=4.5
# VlnPlot_scCustom(bulk, features = c("IL23A"),
#                  pt.size = 0.5,
#                  colors_use = c("Grey", "Red"),
#                  group.by = "celltype",
#                  split.by = "disease",
#                  y.max = 5) +
#   stat_compare_means(method = "anova",
#                      label = "p.signif") +
#   theme(axis.title.x = element_blank(),
#         axis.text.x = element_text(angle = 45))
# cell types on y axis: fig.width=6.5, fig.height=4.5
lungSik <- VlnPlot_scCustom(bulk, features = c("IL23A"),
                 pt.size = 0.5,
                 colors_use = c("Red", "Pink", "Maroon", "Salmon", "Gray"),
                 group.by = "celltype",
                 split.by = "disease") +
stat_compare_means(method = "anova",
                     label = "p.signif",
                     label.y = 11) +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.5)) +
  scale_y_continuous(breaks=seq(0,12,3),
                     limits=c(0, 12)) +
  coord_flip()

# cell types on y axis: fig.width=5.5, fig.height=7
cairo_pdf(filename="/home/williamsdrw/epithelial-IL23/08_final_plots_for_figures/IL23A-Sikkema2023-lungCancer.pdf",
          width=7,
          height=3)
lungSik  
dev.off()
```

### Kong 2023 - Terminal Ileum
```{r k2, fig.width=4.5, fig.height=3}
seurat_objects[[6]]$ref <- "Kong 2023"
seurat_objects[[6]]$source <- "czi_cellxgene"
meta <- seurat_objects[[6]]$biosample_id
meta <- gsub("_", "-", as.character(meta))
seurat_objects[[6]]$orig.ident <- meta
# This dataset was downloaded as epithelial only
# seurat_objects[[6]] <- downsampleSeurat(seurat_objects[[6]], indx)
seurat_objects[[6]] <- convertEnsemblRownames(seurat_objects[[6]])

# generate a seurat object with bulk expression based on cell type, disease status, and donor
bulk <- getBulkExpression(seu = seurat_objects[[6]],
                          cellType = "cell_type",
                          donor = "orig.ident",
                          disease = "disease",
                          diseaseLevels = c("Crohn disease", "normal"))
Idents(bulk) <- "celltype"
bulk <- subset(bulk,
               idents = c("type L enteroendocrine cell",
                          "type EC enteroendocrine cell"),
               invert = T)

# # cell types on x axis: fig.width=6.5, fig.height=4.5
# VlnPlot_scCustom(bulk, features = c("IL23A"),
#                  pt.size = 0.5,
#                  colors_use = c("Grey", "Red"),
#                  group.by = "celltype",
#                  split.by = "disease",
#                  y.max = 3) +
#   stat_compare_means(method = "t.test",
#                      label = "p.signif") +
#   theme(axis.title.x = element_blank(),
#         axis.text.x = element_text(angle = 45))
# cell types on y axis: fig.width=6.5, fig.height=4.5
ileum <- VlnPlot_scCustom(bulk, features = c("IL23A"),
                 pt.size = 0.5,
                 colors_use = c("Red", "Gray"),
                 group.by = "celltype",
                 split.by = "disease",
                 y.max = 4.5) +
  stat_compare_means(method = "t.test",
                     label = "p.signif",
                     label.y = 4.25) +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.5)) +
  coord_flip()

# cell types on y axis: fig.width=5.5, fig.height=7
cairo_pdf(filename="/home/williamsdrw/epithelial-IL23/08_final_plots_for_figures/IL23A-Kong2023-ileum.pdf",
          width=4.5,
          height=3)
ileum  
dev.off()
```

### Tabula Sapiens dataset
```{r tabula sapiens, fig.width=14, fig.height=10}
seurat_objects[[5]]$ref <- "Tabula Sapiens 2022"
seurat_objects[[5]]$source <- "czi_cellxgene"
meta <- seurat_objects[[5]]$donor_id
meta <- gsub("_", "-", as.character(meta))
seurat_objects[[5]]$orig.ident <- meta
meta <- seurat_objects[[5]]$tissue_in_publication
meta <- gsub("_", " ", as.character(meta))
seurat_objects[[5]]$tissue_in_publication <- meta

# This dataset was downloaded as epithelial only
# seurat_objects[[5]] <- downsampleSeurat(seurat_objects[[5]], indx)
seurat_objects[[5]] <- convertEnsemblRownames(seurat_objects[[5]])

# generate a seurat object with bulk expression based on cell type, disease status, and donor
bulk <- getBulkExpression(seu = seurat_objects[[5]],
                          cellType = "cell_type",
                          donor = "orig.ident",
                          disease = "tissue_in_publication", # TS dataset only has normal
                          diseaseLevels = rev(c("Eye",
                                            "Salivary Gland",
                                            "Tongue",
                                            "Trachea",
                                            "Thymus",
                                            "Mammary",
                                            "Heart",
                                            "Lung",
                                            "Large Intestine",
                                            "Small Intestine",
                                            "Liver",
                                            "Pancreas",
                                            "Kidney",
                                            "Bladder",
                                            "Uterus",
                                            "Prostate",
                                            "Vasculature",
                                            "Skin"))) # use the tissue type instead of disease for this dataset
bulk$tissueType <- paste0(bulk$disease, "_", bulk$celltype)
seurat_objects[[5]]$tissueType <- paste0(seurat_objects[[5]]$tissue_in_publication, "_", seurat_objects[[5]]$cell_type)

# set.seed(1105)
# pal <- sample(c(paletteer_d(`"dutchmasters::milkmaid"`),
#                 paletteer_d(`"nord::afternoon_prarie"`),
#                 paletteer_d(`"ggsci::default_aaas"`),
#                 paletteer_d(`"ggsci::default_jama"`)))
# tsVln <- VlnPlot_scCustom(bulk, 
#                    features = c("IL23A"),
#                    pt.size = 0.5,
#                    group.by = "tissueType") +
#     theme(axis.title.y = element_blank(),
#           legend.position = 'none') +
#     coord_flip()
# 
# tsVln2 <- VlnPlot_scCustom(seurat_objects[[5]], 
#                    features = c("IL23A"),
#                    pt.size = 0.5,
#                    group.by = "tissueType") +
#     theme(axis.title.y = element_blank(),
#           legend.position = 'none') +
#     coord_flip()
# 
# tsDot <- DotPlot_scCustom(bulk, 
#                    features = c("IL23A"),
#                    group.by = "tissueType",
#                    colors_use = c("grey", "black")) +
#     theme(axis.title.y = element_blank())
# tsDot2 <- DotPlot_scCustom(seurat_objects[[5]], 
#                    features = c("IL23A"),
#                    group.by = "tissueType",
#                    colors_use = c("grey", "black")) +
#     theme(axis.title.y = element_blank())
# 
# tsDot|tsDot2
# tsVln|tsVln2
# 
# # cell types on y axis: fig.width=6.5, fig.height=4.5
# cairo_pdf(filename="test.pdf",
#           width=3.5,
#           height=6)
# tsVln  
# dev.off()
```

```{r, fig.width=5.5, fig.height=7}
Idents(seurat_objects[[5]]) <- "tissueType"
ts_sub <- subset(seurat_objects[[5]],
                 idents = c("Bladder_bladder urothelial cell",
                            "Eye_conjunctival epithelial cell",
                            "Eye_corneal epithelial cell",
                            "Eye_eye photoreceptor cell",
                            "Eye_keratocyte",
                            "Kidney_kidney epithelial cell",
                            "Large Intestine_enterocyte",
                            "Large Intestine_enterocyte of epithelium of large intestine",
                            "Large Intestine_goblet cell",
                            "Liver_hepatocyte",
                            "Lung_basal cell",
                            "Lung_club cell",
                            "Lung_respiratory goblet cell",
                            "Lung_type II pneumocyte",
                            "Pancreas_pancreatic acinar cell",
                            "Pancreas_pancreatic ductal cell",
                            "Prostate_basal cell of prostate epithelium",
                            "Prostate_club cell",
                            "Prostate_epithelial cell",
                            "Prostate_luminal cell of prostate epithelium",
                            "Salivary Gland_acinar cell of salivary gland",
                            "Salivary Gland_basal cell",
                            "Salivary Gland_duct epithelial cell",
                            "Salivary Gland_salivary gland cell",
                            "Skin_epithelial cell",
                            "Small Intestine_enterocyte of epithelium of small intestine",
                            "Tongue_basal cell",
                            "Tongue_epithelial cell",
                            "Tongue_keratinocyte",
                            "Trachea_basal cell",
                            "Trachea_ciliated cell",
                            "Trachea_secretory cell",
                            "Trachea_tracheal goblet cell",
                            "Uterus_ciliated epithelial cell",
                            "Uterus_epithelial cell",
                            "Uterus_epithelial cell of uterus"))

tsVln3 <- VlnPlot_scCustom(ts_sub, 
                   features = c("IL23A"),
                   pt.size = 0.5,
                   group.by = "tissueType") +
    theme(axis.title.y = element_blank(),
          legend.position = 'none') +
    coord_flip()
bulk2 <- getBulkExpression(seu = ts_sub,
                          cellType = "cell_type",
                          donor = "orig.ident",
                          disease = "tissue_in_publication", # TS dataset only has normal
                          diseaseLevels = rev(c("Eye",
                                            "Salivary Gland",
                                            "Tongue",
                                            "Trachea",
                                            "Thymus",
                                            "Mammary",
                                            "Heart",
                                            "Lung",
                                            "Large Intestine",
                                            "Small Intestine",
                                            "Liver",
                                            "Pancreas",
                                            "Kidney",
                                            "Bladder",
                                            "Uterus",
                                            "Prostate",
                                            "Vasculature",
                                            "Skin"))) # use the tissue type instead of disease for this dataset
bulk2$tissueType <- paste0(bulk2$disease, "_", bulk2$celltype)

tsVln4 <- VlnPlot_scCustom(bulk2, 
                   features = c("IL23A"),
                   pt.size = 0.5,
                   group.by = "tissueType",
                   colors_use = pal,
                   add.noise = F) +
    theme(axis.title.y = element_blank(),
          legend.position = 'none') +
    coord_flip()

# tsDot3 <- DotPlot_scCustom(bulk2, 
#                    features = c("IL23A"),
#                    group.by = "tissueType",
#                    colors_use = c("grey", "black")) +
#     theme(axis.title.y = element_blank())
# tsDot4 <- DotPlot_scCustom(ts_sub, 
#                    features = c("IL23A"),
#                    group.by = "tissueType",
#                    colors_use = c("grey", "black")) +
#     theme(axis.title.y = element_blank())

tsVln4
# tsDot4|tsDot3

# cell types on y axis: fig.width=5.5, fig.height=7
cairo_pdf(filename="/home/williamsdrw/epithelial-IL23/08_final_plots_for_figures/IL23A-tabSapSubset-tsVln4.pdf",
          width=5.5,
          height=7)
tsVln4  
dev.off()
```

# Import non-CZI skin dataset
```{r}
r <- Read10X("/home/williamsdrw/epithelial-IL23/02_raw_data/reynolds21")
r <- CreateSeuratObject(r, project = "Reynolds2021")
```

## Format seurat object (metadata tidying, subset epi)
```{r add meta}
# add meta
meta <- fread("/home/williamsdrw/epithelial-IL23/02_raw_data/reynolds21/metadata.csv")
rownames(meta) <- meta$index
r <- AddMetaData(r, meta)
r$index <- NULL
meta <- r$final_clustering
meta <- gsub("_", "-", as.character(meta))
r$final_clustering <- meta

# subset epi
Idents(r) <- "final_clustering"
idents <- unique(r$final_clustering)[c(1,3,5)]
r <- subset(r, idents = idents)

Idents(r) <- "Status"
r <- RenameIdents(r,
                  "Healthy"="normal")

# re-order disease status
levels(r) <- c(
  "normal",
  "Eczema",
  "Psoriasis"
)
r$Status <- r@active.ident
```

## Standard workflow
```{r norm/scale}
r <- NormalizeData(r) 
r <- FindVariableFeatures(r)
r <- ScaleData(r)
r <- RunPCA(r)
```
## Visualize
```{r visualize, fig.width=4.5, fig.height=3}
# generate a seurat object with bulk expression based on cell type, disease status, and donor
bulk <- getBulkExpression(seu = r,
                          cellType = "final_clustering",
                          donor = "donor_id",
                          disease = "Status",
                          diseaseLevels = c("Psoriasis", "Eczema", "normal"))

# cell types on x axis: fig.width=6.5, fig.height=4.5
skin <- VlnPlot_scCustom(bulk, features = c("IL23A"),
                 pt.size = 0.5,
                 colors_use = c("Red", "Maroon", "Gray"),
                 group.by = "celltype", 
                 split.by = "disease",
                 y.max = 8) + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.5)) +
    scale_y_continuous(breaks=seq(0,8,2),
                     limits=c(0, 8)) +
  coord_flip()

skindf$split<- as.factor(skindf$split)

skindf <- skin$data 
stat <- skindf %>%
  group_by(ident) %>%
  t_test(IL23A ~ split, ref.group = "normal")

# cell types on y axis: fig.width=5.5, fig.height=7
cairo_pdf(filename="/home/williamsdrw/epithelial-IL23/08_final_plots_for_figures/IL23A-Reynolds2021-skin.pdf",
          width=4.5,
          height=3)
skin  
dev.off()
```



# Save processed objects
```{r save}
# # combine all seurat files
# all <- merge(x=NIH, y=c(y, c, k1, s, t, k2))
# save seurat obj with meta
baseDir <- "~/epithelial-IL23/04_data_objects/03_Analysis_objects/"
saveRDS(seurat_objects, file = paste0(baseDir, Sys.Date(), "_05_02-ExternalOnly_czi.RDS"))
saveRDS(r, file = paste0(baseDir, Sys.Date(), "_05_02-ExternalOnly_skin.RDS"))
```

```{r session info}
sessionInfo()
```