---
title: "Unify Metadata and Merge Datasets"
author: "Drake Williams"
date: "`r BiocStyle::doc_date()`"
output: html_document
knit: (function(inputFile, encoding) { 
    rmarkdown::render(
        inputFile, encoding = encoding, 
        output_file = file.path(
            dirname(inputFile), paste0('01.5_Metadata_',Sys.Date(),'.html'))) 
                })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = F,
  message = F, 
  out.width = "100%",
  fig.align = "center",
  time_it = T
)
options(width = 1200)
```

# Introduction

The objective of this notebook is to load datasets from CZI CELLxGENE and change ENSEMBL gene names to gene symbols

## Load packages

```{r load packages}
# load packages
library(pacman)
p_load(Seurat, 
       scCustomize, 
       dplyr, 
       org.Hs.eg.db, 
       ggpubr)
```
## Import all CZI datasets
```{r generate czi seurat list, cache=TRUE}
# Data came in the format of an individual RDS
# Function to load all data in the folder into a list of seurat objects
read_rds_files <- function(folder) {
  # Get a list of all RDS files in the folder
  rds_files <- list.files(folder, pattern = "\\.rds$")

  # Add the path to each RDS file
  rds_files <- file.path(folder, rds_files)

  # Read each RDS file and create a Seurat object
  seurat_objects <- lapply(rds_files, readRDS)

  # Return a list of Seurat objects, where the name of each object is the name of the file
  return(setNames(seurat_objects, rds_files))
}

folder <- "/home/williamsdrw/epithelial-IL23/02_raw_data/czi/"
seurat_objects <- read_rds_files(folder)
```

## Modify metadata and trim CZI datasets
```{r}
# modify metadata
seurat_objects[[1]]$ref <- "Yoshida 2022"
seurat_objects[[1]]$source <- "czi_cellxgene"
seurat_objects[[1]]$orig.ident <- seurat_objects[[1]]$sample_id
# extract epithelial cells
Idents(seurat_objects[[1]]) <- "Cell_type_annotation_level1"
y <- subset(seurat_objects[[1]], idents = "Epi")
```

```{r}
seurat_objects[[2]]$ref <- "Chan 2021"
seurat_objects[[2]]$source <- "czi_cellxgene"
seurat_objects[[2]]$orig.ident <- seurat_objects[[2]]$donor_id
 # extract epithelial cells
Idents(seurat_objects[[2]]) <- "cell_type"
c <- subset(seurat_objects[[2]], idents = c("ionocyte", "hepatocyte", "neuroendocrine cell"), invert = T)
```

```{r}
seurat_objects[[3]]$ref <- "Kong 2023"
seurat_objects[[3]]$source <- "czi_cellxgene"
seurat_objects[[3]]$orig.ident <- seurat_objects[[3]]$biosample_id
# This dataset was downloaded as epithelial only
k1 <- seurat_objects[[3]]
```

```{r}
seurat_objects[[4]]$ref <- "Sikkema 2022"
seurat_objects[[4]]$source <- "czi_cellxgene"
seurat_objects[[4]]$orig.ident <- seurat_objects[[4]]$sample
# extract epithelial cells
Idents(seurat_objects[[4]]) <- "ann_coarse"
s <- subset(seurat_objects[[4]], idents = c("Epithelial cell"))
```

```{r}
seurat_objects[[5]]$ref <- "Tabula Sapiens 2022"
seurat_objects[[5]]$source <- "czi_cellxgene"
seurat_objects[[5]]$orig.ident <- seurat_objects[[5]]$donor_id
# This dataset was downloaded as epithelial only
t <- seurat_objects[[5]]
```

```{r, fig.width=6.5, fig.height=4.5}
seurat_objects[[6]]$ref <- "Kong 2023"
seurat_objects[[6]]$source <- "czi_cellxgene"
seurat_objects[[6]]$orig.ident <- seurat_objects[[6]]$biosample_id
# This dataset was downloaded as epithelial only
k2 <- seurat_objects[[6]]



# cell types on x axis: fig.width=6.5, fig.height=4.5
# cell types on y axis: 
VlnPlot_scCustom(bulk, features = c("IL23A"),
                 pt.size = 0.5,
                 colors_use = c("Grey", "Red"),
                 group.by = "celltype", 
                 split.by = "disease",
                 y.max = 2.5) + 
  stat_compare_means(method = "t.test",
                     label = "p.signif") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45))

# cell types on x axis: fig.width=6.5, fig.height=4.5
VlnPlot_scCustom(bulk, features = c("IL23A"),
                 pt.size = 0.5,
                 colors_use = c("Grey", "Red"),
                 group.by = "celltype", 
                 split.by = "disease",
                 y.max = 2.5) + 
  stat_compare_means(method = "t.test",
                     label = "p.signif") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45))
# cell types on y axis: fig.width=6.5, fig.height=4.5
VlnPlot_scCustom(bulk, features = c("IL23A"),
                 pt.size = 0.5,
                 colors_use = c("Grey", "Red"),
                 group.by = "celltype", 
                 split.by = "disease",
                 y.max = 2.5) + 
  stat_compare_means(method = "t.test",
                     label = "p.signif") +
  theme(axis.title.y = element_blank()) + 
  coord_flip()
```

## Downsample datasets to 20k cells
The datasets from CZI also have ENSEMBL gene names instead of gene symbols. For each of the datasets, feature names are converted to the appropriate symbol after downsampling
```{r downsample objects}
source("~/epithelial-IL23/01_scripts/analysisHelpers.R")
indx <- 20000

y <- downsampleSeurat(y, indx)
y <- convertEnsemblRownames(y)
c <- downsampleSeurat(c, indx)
c <- convertEnsemblRownames(c)
k1 <- downsampleSeurat(k1, indx)
k1 <- convertEnsemblRownames(k1)
s <- downsampleSeurat(s, indx)
s <- convertEnsemblRownames(s)
t <- downsampleSeurat(t, indx)
t <- convertEnsemblRownames(t)
k2 <- downsampleSeurat(k2, indx)
k2 <- convertEnsemblRownames(k2)
```

```{r clear memory}
seurat_objects <- NULL
gc()
```

## Merge all datasets and save for the next step
```{r combine objects and save}
# combine all seurat files
all <- merge(x=NIH, y=c(y, c, k1, s, t, k2))
# save seurat obj with meta
baseDir <- "~/epithelial-IL23/04_data_objects/01_SoupX_objects/"
saveRDS(all, file = paste0(baseDir, Sys.Date(), "_015-allWithMeta.RDS"))
```

```{r session info}
sessionInfo()
```

